---
title: "Exploratory Data Analysis"
author: "Mauro Lepore"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 6
vignette: >
  %\VignetteIndexEntry{Exploratory Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

### Introduction

This vignette will show you how to explore your data systematically, using visualization and transformation. The goal is to use your data and some research questions to learn something that you can then use to refine your questions. You should repeat the process until you are happy with what you learned.

We will use data from Barro Colorado Island (DOI: https://doi.org/10.5479/data.bci.20130603), and functions from the packages **[forestr](xxxaddlink)**, **[dplyr](xxxaddlink)** and **[ggplot2](xxxaddlink)**.

```{r packages}
# To install packages see ?install.packages and ?devtools::install_github
library(bci)      # to access data of forest dynamics from BCI
library(forestr)  # to analyse forest dynamics
library(ggplot2)  # to visualize data
library(dplyr)    # to transform data
```

#### R for Data Science

The ideas presented here come from [Exploratory Data Analysis](http://r4ds.had.co.nz/exploratory-data-analysis.html), in [R for Data Science](http://r4ds.had.co.nz/), by Garrett Grolemund and Hadley Wickham. Some text from that book section is literally reproduced here _like this_ or

> like this,

```R
but not like this (compared to the text above, this one has different indentation and fonttype).
```

For example, the following definitions are literally reproduced:

> A __variable__ is a quantity, quality, or property that you can measure. 

> A __value__ is the state of a variable when you measure it. The value of a
    variable may change from measurement to measurement.
  
> An __observation__ is a set of measurements made under similar conditions
    (you usually make all of the measurements in an observation at the same 
    time and on the same object). An observation will contain several values, 
    each associated with a different variable. I'll sometimes refer to 
    an observation as a data point.

> __Tabular data__ is a set of values, each associated with a variable and an 
    observation. Tabular data is _tidy_ if each value is placed in its own
    "cell", each variable in its own column, and each observation in its own 
    row. 



### Questions

To start exploring your data, you can generally ask:

> What type of variation occurs within my variables?

> What type of covariation occurs between my variables?


### Variation

> __Variation__ is the tendency of the values of a variable to change from measurement to measurement. 

Every variable varies with a particular pattern, and that pattern may be insightful. To understand the variation pattern of a variable we can visualize the distribution of the variables' values. The best way to visualize a variable's distribution depends on whether the variable is categorical or continuous.

#### Categorical variables

> To examine the distribution of a categorical variable, use a bar chart

```{r}
# xxx replace with bci data
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

xxxcont.

#### Wrap up

To do all of the above quickly, we can use package __skimr__ (https://github.com/ropenscilabs/skimr).

```{r}
library(skimr)
```

```{r}
smry <- skim(diamonds)
smry %>% 
  dplyr::filter(stat == "hist") %>% 
  as.matrix()  # Only needed in Windows for histograms (https://goo.gl/S8MaZW)
```


#### Continuous variables




