---
title: "Growth variation through time"
subtitle: "Adapted by Mauro Lepore from the original tutorial by Richard Condit (http://ctfs.si.edu/Public/CTFSRPackage/files/tutorials/GrowthChange/)"
author: "Richard Condit"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"growth variation through time"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

# Hypotheses

The question addressed here is how growth has changed with time, in particular how species differ in growth changes. There are two alternative models described:

* Growth rate changing linearly with time
* Growth rate varying from census to census, but not following a consistent change

# Packages

The functions we use here come with the [ctfs](https://forestgeo.github.io/ctfs/) and [lme4](https://github.com/lme4/lme4/) packages, and the data comes in the [bci](https://forestgeo.github.io/bci/) package.

```{r}
# installation instructions at https://github.com/forestgeo/ctfs
library(ctfs)
# installation instructions at https://github.com/lme4/lme4/
library(lme4)
# installation instructions at https://forestgeo.github.io/bci/
library(bci)
```

# Formatting data






# References

[Repository](https://dx.doi.org/10.5479/data.bci.20130603) of the Barro Colorado Forest Census Plot Data (version 2012).

[Description](https://repository.si.edu/bitstream/handle/10088/20925/RoutputFull.pdf?sequence=1&isAllowed=y) of the Barro Colorado Forest Census Plot Data.



